VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-22 doc/publications/journal2014/paper.tex
couple of changes

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and {\sc ADAM} both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
We attribute the 30\% speedup over {\sc ADAM} (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas {\sc ADAM} creates dense vectors, which are less memory efficient. %
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
Despite this, both are an order of magnitude slower than \variantSpark{} and {\sc ADAM}, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
{\sc Admixture} is also slower with a runtime of just over 8 minutes. 
Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863
Hadoop writes the entire output of each \kMeans{} iteration to disk. This feature was designed to ensure scalability but here substantially increases the runtime as disk IO becomes the limiting factor for this small dataset (chromosome 22).
Python and R store the output from each iteration in memory thereby eliminating the IO bottleneck but without being able to utilise the disk they are limited to datasets that fit into memory.  
{\sc Spark} overcomes both limitations by performing in-memory caching for each executor, which eliminates the disk IO bottleneck on smaller datasets, while maintaining scalability to larger datasets where the data actively processed on executors is cashed in memory.

@@

\subsection*{VariantSpark Implementation}
To cluster the individuals in a VCF file, we initially need to pre-process the variants to feature vectors.
In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD). RDDs allow us to process the files, line-by-line, in parallel. We parse each line as tab-separated values, and Spark's `zip' each value with it's heading (Sample name) from the VCF file, resulting in key-value pairs (KVPs) for each line see Figure 2 (Spark step 1). 
We then `zip' each pair with a unique sequential index, which later-on will serve as identifier for each allele (and keep the genome loci order), resulting in a KVP array (step 2).
We use `flatMap' to transpose the data structure to facilitate the grouping of variants by individuals.
For the \kMeans{} clustering, we convert the VCF genotypes (i.e. `0\textbar{}1') to doubles using the Hamming distance to the reference, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant. 
To remove uninformative KVPs,  we filter out any alleles that have no variant. 
We perform this `map' operation inside the previous `flatMap' (step 3).
We can now group these KVPs by their key (the individual ID) and save them to sparse vectors (step 4).

>>>>

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and {\sc ADAM} both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
We attribute the 30\% speedup over {\sc ADAM} (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas {\sc ADAM} creates dense vectors, which are less memory efficient. %
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
Despite this, both are an order of magnitude slower than \variantSpark{} and {\sc ADAM}, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
{\sc Admixture} is also slower with a runtime of just over 8 minutes. 
Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863
Hadoop writes the entire output of each \kMeans{} iteration to disk. This feature was designed to ensure scalability but here substantially increases the runtime as disk IO becomes the limiting factor for this small dataset (chromosome 22).
Python and R store the output from each iteration in memory, thereby eliminating the IO bottleneck, but without being able to utilise the disk they are limited to datasets that fit into memory.
{\sc Spark} overcomes both limitations by performing in-memory caching for each executor, which eliminates the disk IO bottleneck on smaller datasets, while maintaining scalability to larger datasets by spilling data that exceeds memory availability to disk.

@@

\subsection*{VariantSpark Implementation}
To cluster the individuals in a VCF file, we initially need to pre-process the variants to feature vectors.
In \variantSpark{} we read in the VCF files as text files to a Resilient Distributed Dataset (RDD). RDDs allow us to process the files, line-by-line, in parallel. We parse each line as tab-separated values, and with Spark's `zip' function, `zip' each value with it's heading (Sample name) from the VCF file, resulting in key-value pairs (KVPs) for each line see Figure 2 (Spark step 1). 
We then `zip' each pair with a unique sequential index, which later-on will serve as identifier for each allele (and keep the genome loci order), resulting in a KVP array (step 2).
We use `flatMap' to transpose the data structure to facilitate the grouping of variants by individuals.
For the \kMeans{} clustering, we convert the VCF genotypes (i.e. `0\textbar{}1') to doubles using the Hamming distance to the reference, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant. 
To remove uninformative KVPs, we filter out any alleles that have no variant. 
We perform this `map' operation inside the previous `flatMap' (step 3).
We can now group these KVPs by their key (the individual ID) and save them to sparse vectors (step 4).
